\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[hidelinks]{hyperref}
\usepackage{url}
\usepackage{listings}
\usepackage{caption}
\usepackage{multirow}

\definecolor{codebg}{rgb}{0.97,0.97,0.97}
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  frame=single,
  framerule=0pt,
  showstringspaces=false,
  breaklines=true,
  language=C,
  keywordstyle=\color{blue!60!black},
  commentstyle=\color{green!50!black}\itshape,
  stringstyle=\color{red!60!black},
}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Zq}{\Z_q}
\newcommand{\Rq}{R_q}
\newcommand{\NTT}{\mathrm{NTT}}
\newcommand{\iNTT}{\mathrm{NTT}^{-1}}
\newcommand{\Power}{\mathrm{Power2Round}}
\newcommand{\HighBits}{\mathrm{HighBits}}
\newcommand{\LowBits}{\mathrm{LowBits}}
\newcommand{\UseHint}{\mathrm{UseHint}}
\newcommand{\MakeHint}{\mathrm{MakeHint}}
\newcommand{\ExpandA}{\mathrm{ExpandA}}
\newcommand{\SHAKE}{\mathrm{SHAKE}}
\newcommand{\bigO}{\mathcal{O}}

\title{\Large\bfseries Specializing ML-DSA-44 Verification for Secure Boot on Cortex-M4 \\
       \large A Didactic Walk-through of Ten Progressive Optimizations}
\date{2026-05-11}
\author{}

\begin{document}
\maketitle

\begin{abstract}
We present a custom verifier for ML-DSA-44 (NIST FIPS 204) targeted at
secure-boot deployments on the ARM Cortex-M4. Starting from
\texttt{pqm4}'s \texttt{m4f} backend --- the fastest publicly
available Cortex-M4 verifier we benchmarked, beating both
\texttt{mldsa-native}'s portable C90 and the round-3 Dilithium2
implementation in \texttt{FasterKyberDilithiumM4} on this metric ---
we apply ten layered optimizations and reduce per-verify cost from
$1{,}261{,}555$ to $241{,}756$ instructions retired on the
Cortex-M4 ISA --- a $\mathbf{5.22\times}$ speedup. The optimizations exploit
two observations that hold for secure boot but not for general ML-DSA
verification:
(i)~the public key is fixed at provisioning time, so any quantity that
depends only on $\textsf{pk}$ may be precomputed and stored in flash;
(ii)~all verifier inputs are public, so the constant-time discipline of
\texttt{pqm4} is no longer load-bearing.

The verifier is \emph{not} a drop-in replacement for a general ML-DSA-44
verify routine: it is hard-specialized to the provisioned public key
and rejects any input with non-empty context. On the secure-boot input
subset (ML-DSA-44, the provisioned $\textsf{pk}$, $|\mathrm{ctx}|=0$),
nine of the ten variants are accept/reject-equivalent to FIPS 204
Algorithm 3 \emph{by construction} (every transformation either
precomputes a deterministic function of $\textsf{pk}$ or refactors C
code without changing its behaviour). One variant ($\textsf{v9}$,
which removes a polynomial reduction we conjecture to be redundant
under our pre-invNTT bounds) has an open soundness question we have
not closed by worst-case analysis; we verify it empirically on the
test vector $\texttt{mldsa-native}$ produces. We give both high-level
intuition and the precise transformations applied at each step, with
measurements at every checkpoint reproducible from a small QEMU
harness.\smallskip

\noindent\textbf{Headline:} ML-DSA-44 signature verification on Cortex-M4
becomes $\sim$34$\times$ cheaper than pure-C P-256 ECDSA verify and
$\sim$51$\times$ cheaper than the hand-asm \texttt{p256-m}.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The migration from elliptic-curve cryptography (ECC) to post-quantum
cryptography is now operational policy: NIST published the first three
post-quantum standards in August~2024, including FIPS~204 (ML-DSA, a
digital signature algorithm based on the Dilithium~\cite{dilithium} module
lattice construction). Secure-boot is among the first deployment domains
where the new standard matters: the public verification key is provisioned
into a device at manufacture, then the bootloader verifies a firmware
signature on every boot until end-of-life. Verifier cost is paid every
boot; key size is paid once.

ML-DSA's verify path is dominated by polynomial arithmetic over the ring
$\Rq = \Zq[X]/(X^{256}+1)$, $q = 2^{23} - 2^{13} + 1 = 8{,}380{,}417$, with
heavy use of the Number-Theoretic Transform (NTT). Compared to a
P-256 ECDSA verify on the same Cortex-M4, an off-the-shelf ML-DSA-44
verify in C is already $\sim$3--4$\times$ cheaper; with the hand-tuned
\texttt{pqm4}~\cite{pqm4} M4 asm it becomes $\sim$6$\times$ cheaper. This
paper asks how much further the cost can fall when one exploits the
\emph{fixed-public-key, all-public-data} setting that defines secure boot.

The contribution is twofold. First, we measure ten ML-DSA-44 verifier
variants on a single bare-metal Cortex-M4 testbench, with reproducible
instruction-retired counts. Second, and primarily, we deliver a didactic
exposition: each optimization is presented with its high-level
justification, the precise FIPS~204 step it touches, and its measured
impact, so a practitioner can decide which transformations apply to their
threat model and which do not. We are explicit about what we do
\emph{not} skip --- the FIPS-204-mandated checks on $z$'s norm and the
hint encoding remain in place --- because for secure boot the
correctness-of-verify property is the entire load-bearing assumption.

\section{Background}
\label{sec:background}

This section gives just enough lattice-cryptography to follow the
verification anatomy in Section~\ref{sec:verify-anatomy}. Readers fluent
in module-LWE and the Dilithium structure may skim.

\subsection{Module-LWE and the Ring \texorpdfstring{$\Rq$}{Rq}}

Let $q$ be a prime and $n$ a power of two. Define the
\emph{negacyclic} ring
\[
  \Rq \;:=\; \Zq[X] / (X^n + 1).
\]
Elements of $\Rq$ are polynomials of degree $<n$ with coefficients in
$\Zq$; multiplication is polynomial multiplication modulo $X^n + 1$, which
under our parameters ($n=256$, $q = 2^{23} - 2^{13} + 1$) admits a fast
$n\log n$ Number-Theoretic Transform.

The \emph{module} variant of the Learning-With-Errors problem (M-LWE)
fixes ranks $k,\ell \in \N$ and asks: distinguish
\[
  (\mathbf{A},\,\mathbf{A}\mathbf{s} + \mathbf{e})
  \quad\text{from}\quad
  (\mathbf{A},\,\mathbf{u})
\]
where $\mathbf{A} \in \Rq^{k\times\ell}$ is uniform, $\mathbf{s} \in
\Rq^{\ell}$ and $\mathbf{e} \in \Rq^{k}$ are short (coefficients in
$\{-\eta,\dots,\eta\}$ for a small $\eta$), and $\mathbf{u}$ is uniform in
$\Rq^{k}$. Both Kyber (now ML-KEM) and Dilithium (now ML-DSA) reduce their
security to M-LWE and its sibling M-SIS.

\subsection{Kyber (ML-KEM) at a Glance}

ML-KEM (FIPS 203) is a key-encapsulation mechanism whose IND-CPA core is a
ring-flavoured Regev-style PKE: the public key contains an M-LWE sample
$\mathbf{t} = \mathbf{A}\mathbf{s} + \mathbf{e}$; encapsulation draws short
$(\mathbf{r}, \mathbf{e}_1, e_2)$ and computes a ciphertext of the form
$(\mathbf{u},v) = (\mathbf{A}^\top\mathbf{r} + \mathbf{e}_1,\;
\mathbf{t}^\top\mathbf{r} + e_2 + \delta(m))$
where the message $m \in \{0,1\}^n$ is lifted by $\delta(m)_i := m_i
\cdot ((q+1)/2)$ to put each bit either at $0$ or at the integer
closest to $q/2$ in $\Zq$. Decapsulation recovers $m_i$ by checking
whether $(v - \mathbf{s}^\top \mathbf{u})_i$ is closer to $0$ or to
$(q+1)/2$, which succeeds whenever the error term
$\mathbf{e}^\top\mathbf{r} - \mathbf{s}^\top\mathbf{e}_1 + e_2 - \mathbf{s}^\top \mathbf{e}$
has every coefficient of magnitude $< q/4$. Kyber additionally
\emph{compresses} the ciphertext and public key by truncating each
coefficient to a few high-order bits (we elide that step here for the
sketch), then wraps the IND-CPA scheme in the Fujisaki--Okamoto
transform to achieve IND-CCA. We do not optimize Kyber in this work,
but the $\mathbf{A}\mathbf{s} + \mathbf{e}$ pattern and the same
underlying ring recur in Dilithium below.

\subsection{Dilithium (ML-DSA) Construction}
\label{sec:dilithium}

ML-DSA is a \emph{Fiat--Shamir-with-aborts} signature: a $\Sigma$-protocol
proving knowledge of a short M-LWE secret, made non-interactive by
deriving the challenge via a random oracle ($\SHAKE$-256). The
construction's three signature levels are parametrized as in
Table~\ref{tab:mldsa-params}.

\begin{table}[h]
\centering
\caption{ML-DSA parameter sets (FIPS 204, \S4 Table~1).}
\label{tab:mldsa-params}
\begin{tabular}{l c c c c c c c c c c}
\toprule
            & $q$       & $n$  & $d$ & $\tau$ & $\lambda$ & $\gamma_1$ & $\gamma_2$       & $(k,\ell)$ & $\eta$ & $\omega$ \\
\midrule
ML-DSA-44   & $2^{23}{-}2^{13}{+}1$ & 256 & 13 & 39 & 128 & $2^{17}$  & $(q{-}1)/88$ & $(4,4)$ & 2 & 80 \\
ML-DSA-65   & ,, & 256 & 13 & 49 & 192 & $2^{19}$  & $(q{-}1)/32$ & $(6,5)$ & 4 & 55 \\
ML-DSA-87   & ,, & 256 & 13 & 60 & 256 & $2^{19}$  & $(q{-}1)/32$ & $(8,7)$ & 2 & 75 \\
\bottomrule
\end{tabular}
\end{table}

For ML-DSA-44 a public key occupies $32 + k \cdot 320 = 1{,}312$~bytes
(rho plus the high-order $t_1$ of $\mathbf{t} = \mathbf{A}\mathbf{s}_1 +
\mathbf{s}_2$ truncated by $\Power$) and a signature occupies
$32 + \ell \cdot 576 + (\omega + k) = 2{,}420$~bytes (the challenge hash
$\tilde c$, the masked response $\mathbf{z}$, and the hint $\mathbf{h}$).

\subsection{The Number-Theoretic Transform}

Because $q-1 = 2^{23} - 2^{13}$ is divisible by $2n = 512$, $\Zq$ contains
a primitive $2n$-th root of unity $\zeta$ (with $\zeta = 1753$ being the
canonical choice in Dilithium). The \emph{negacyclic} NTT evaluates a
polynomial $a(X) \in \Rq$ at the $n$ odd powers of $\zeta$:
\[
  \widehat{a}_i \;=\; a(\zeta^{2i+1}) \pmod{q}, \quad i = 0,\dots,n-1.
\]
Then $a \cdot b$ in $\Rq$ equals $\iNTT(\widehat{a} \odot \widehat{b})$
where $\odot$ is the coefficient-wise product. The forward and inverse
transforms run in $\bigO(n \log n)$ ring operations via the
Cooley--Tukey/Gentleman--Sande structure, so the overall cost of a ring
multiplication drops from $\bigO(n^2)$ schoolbook to $\bigO(n \log n)$
butterflies plus $n$ Montgomery pointwise multiplies. The \texttt{pqm4}
\texttt{m4f} backend implements the entire transform in hand-tuned
Thumb-2 assembly, scheduling loads and stores to overlap with the M4's
multi-cycle 32$\times$32 multiplies; we treat it as a black box throughout.

\subsection{Montgomery Arithmetic}

Multiplications modulo $q$ avoid an expensive division by working in
\emph{Montgomery form}: integers are represented by $\bar a := a R \bmod
q$ for a fixed $R$ coprime to $q$. We use $R = 2^{32}$ throughout. The
\emph{Montgomery reduction} of $T \in [0, qR)$ returns
\[
  \mathrm{Redc}(T) \;:=\; (T + m\,q)/R \;\equiv\; T R^{-1} \pmod q,
  \qquad m \;:=\; T \cdot q' \bmod R,
\]
where $q'$ is the unique element of $\Z_R$ satisfying $q\,q' \equiv -1
\pmod R$ (the choice of sign makes $T + m\,q$ divisible by $R$ so the
quotient is exact). The output sits in $[0, 2q)$, which a single
conditional subtraction (or no subtraction at all, if downstream code
admits a slightly-larger input range) brings into $[0, q)$. For Dilithium
$q = 8{,}380{,}417$ and $R = 2^{32}$, pqm4 precomputes $q' = $
\texttt{0xFC7FDFFF} = $-q^{-1} \bmod 2^{32}$.

The \emph{Montgomery multiplication} of $\bar a, \bar b \in [0, q)$ is
$\mathrm{Mul}(\bar a, \bar b) := \mathrm{Redc}(\bar a \cdot \bar b)$,
which yields $\overline{ab} = (ab)R \bmod q$ --- i.e.\ the product
stays in Montgomery form. In Thumb-2 the M4 implementation is three
multiplies and a 64-bit add: \texttt{SMULL} of $\bar a \cdot \bar b$
producing a 64-bit $(T_{\text{lo}}, T_{\text{hi}})$ in a register pair,
\texttt{MUL} of $T_{\text{lo}} \cdot q'$ giving $m$, then \texttt{SMLAL}
which accumulates $m \cdot q$ into the same register pair so that
$T_{\text{hi}}$ already holds $\mathrm{Redc}(T)$ on output. On Cortex-M4
\texttt{SMULL} and \texttt{SMLAL} are each 3--5 cycles and \texttt{MUL}
is 1 cycle, so a full Montgomery multiplication retires in
$\sim$$8$--$11$~cycles depending on operand magnitudes. Throughout this
paper, ``in NTT domain'' should be read as ``in NTT-Montgomery domain'',
matching the convention pqm4 uses.

\section{ML-DSA-44 Verification, Anatomically}
\label{sec:verify-anatomy}

We rephrase FIPS~204's ML-DSA.Verify (Algorithm~3) and
ML-DSA.Verify\_internal (Algorithm~8) jointly as Algorithm~\ref{alg:verify}
so that one walk-through covers both the public-API formatting layer and
the lattice core. The comments tag each step with the implementation
phase whose cost we will later measure and attack.

\begin{algorithm}[h]
\caption{ML-DSA verification (combined view of FIPS 204 Alg.~3 and Alg.~8).}
\label{alg:verify}
\begin{algorithmic}[1]
\Require Public key $\textsf{pk} = (\rho, t_1)$, message $M$, context string $\mathrm{ctx}$
with $|\mathrm{ctx}| \le 255$, signature $\sigma = (\tilde c, \mathbf{z}, \mathbf{h})$.
\Ensure $\top$ if $\sigma$ is a valid signature on $M$ under $\textsf{pk}$, else $\bot$.
\State $(\tilde c, \mathbf{z}, \mathbf{h}) \gets$ \Call{UnpackSig}{$\sigma$} \Comment{18-bit unpack of $\mathbf{z}$; sparse hint decode; fail-on-malformed}
\If{$\lVert \mathbf{z} \rVert_\infty \ge \gamma_1 - \beta$} \Return $\bot$ \EndIf \Comment{spec-mandated norm validation}
\State $\mathbf{A} \gets \ExpandA(\rho) \in \Rq^{k\times\ell}$ \Comment{$k\ell$ \textbf{rejection-sampled} polynomials, output in NTT domain}
\State $tr \gets \SHAKE_{256}(\textsf{pk}, 64)$ \Comment{64-byte hash of the bit-packed public key}
\State $M' \gets \mathtt{0x00} \,\|\, \mathrm{IntegerToBytes}(|\mathrm{ctx}|, 1) \,\|\, \mathrm{ctx} \,\|\, M$ \Comment{pure ML-DSA prefix}
\State $\mu \gets \SHAKE_{256}(tr \,\|\, M',\, 64)$
\State $c \gets$ \Call{SampleInBall}{$\tilde c$} \Comment{$\tau$-sparse $\{-1,0,+1\}$ polynomial via a SHAKE-256-seeded shuffle}
\State $\hat c \gets \NTT(c)$; \quad $\hat{\mathbf{z}} \gets \NTT(\mathbf{z})$
\State $\hat{\mathbf{w}}' \gets \mathbf{A} \cdot \hat{\mathbf{z}} \;-\; \hat c \cdot \NTT(2^d \cdot t_1)$ \Comment{matrix--vector and one scaling, in NTT domain}
\State $\mathbf{w}'_{\text{approx}} \gets \UseHint(\mathbf{h}, \iNTT(\hat{\mathbf{w}}'))$ \Comment{reconstruct $\HighBits$ with hint correction}
\State $\tilde c' \gets \SHAKE_{256}(\mu \,\|\, \mathrm{w1Encode}(\mathbf{w}'_{\text{approx}}),\, 2\lambda/8)$
\State \Return $\tilde c' \overset{?}{=} \tilde c$
\end{algorithmic}
\end{algorithm}

\paragraph{Where the cost goes.} Rather than estimate per-phase shares
from instruction analysis, we report the cost of each implementation
phase as the \emph{measured difference} between two of our verifier
variants (Section~\ref{sec:opts}). Each variant either keeps or removes
exactly one phase, so the delta is the cost of that phase in the
baseline. The largest contributor is $\ExpandA$, which generates the
matrix $\mathbf{A}$ by rejection-sampling sixteen polynomials from
SHAKE-128 streams keyed by $\rho \,\|\, j \,\|\, i$:

\begin{table}[ht]
\centering
\caption{Cost of each verify phase that we are able to switch off in
\emph{some} $\textsf{sb-mldsa44-v}N$ variant, measured directly as the
per-verify instruction-retired delta between consecutive variants. Total
verify cost in pqm4 \texttt{m4f} (\textsf{v1}) is $1{,}261{,}555$
instructions; cumulative removals account for $1{,}019{,}799$, leaving
$241{,}756$ in \textsf{v10}. Phases not represented here (the four NTTs
on $\mathbf{z}$, the inverse NTT, the matrix--vector pointwise-Mont
multiplications, the $\mu$ and $\tilde c'$ Keccak permutations not tied
to $\textsf{pk}$, and the spec-mandated unpacks and checks) remain on
the critical path.}
\label{tab:cost-breakdown}
\begin{tabular}{l l r}
\toprule
Phase removed                                       & Delta source     & Insns retired \\
\midrule
$\ExpandA(\rho)$ (16 SHAKE-128 rejection-sampling streams) & \textsf{v1}$-$\textsf{v2} & $851{,}263$ \\
$\mathrm{unpack}(t_1) + \mathrm{shiftL}_{2^d} + \NTT$ on $t_1$ & \textsf{v2}$-$\textsf{v3} & $42{,}273$ \\
$\SHAKE_{256}(\textsf{pk}, 64) \to tr$              & \textsf{v3}$-$\textsf{v4}     & $99{,}599$ \\
zero-init + sub-rate XOR of $tr \,\|\, \mathtt{0x00\,0x00}$ & \textsf{v4}$-$\textsf{v5}     & $476$ \\
$K$ redundant walks of the hint encoding             & \textsf{v5}$-$\textsf{v6}     & $2{,}484$ \\
three-pass $\mathrm{caddq}\to\UseHint\to\mathrm{w1Encode}$ & \textsf{v6}$-$\textsf{v7}     & $8{,}901$ \\
separate unpack/chknorm scans over $\mathbf{z}$      & \textsf{v7}$-$\textsf{v8}     & $8{,}771$ \\
$\mathrm{poly\_reduce}$ before $\iNTT$ (redundant)  & \textsf{v8}$-$\textsf{v9}     & $5{,}412$ \\
inc-API plumbing around $\tilde c'$ Keccak           & \textsf{v9}$-$\textsf{v10}    & $620$ \\
\midrule
\textbf{cumulative removal}                          & \textsf{v1}$-$\textsf{v10}    & $\mathbf{1{,}019{,}799}$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Observation.} $\ExpandA$ alone is $851{,}263 / 1{,}261{,}555 =
67.5\%$ of the entire pqm4 \texttt{m4f} verify --- which is precisely
where the biggest payoff hides for a fixed-$\textsf{pk}$ deployment.

\section{The Secure-Boot Threat Model}
\label{sec:secure-boot}

We adopt the following operational assumptions, all standard for ROM-based
secure boot:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Fixed public key.} A single $\textsf{pk}$ is provisioned
        at manufacturing time and is the only key the verifier ever
        consults. Any quantity that is a deterministic function of
        $\textsf{pk}$ may be computed once at provisioning, stored in
        flash, and read on every boot.
  \item \textbf{All inputs are public.} The bootloader does not handle
        secret data during verification --- the firmware image, the
        signature, and the public key are all attacker-visible. The
        constant-time discipline that protects ML-DSA \emph{signing} from
        side-channel attacks is therefore unnecessary on the verify side.
  \item \textbf{Conformant on the secure-boot input subset.} Define
        the secure-boot input subset $\mathcal{I}_{\mathrm{sb}}$ as
        the set of $(\textsf{pk}, M, \mathrm{ctx}, \sigma)$ tuples
        where $\textsf{pk}$ equals the provisioned key and
        $|\mathrm{ctx}| = 0$. The verifier reproduces the
        accept/reject decision of FIPS~204 Algorithm~3 on every input
        in $\mathcal{I}_{\mathrm{sb}}$; on inputs outside this subset
        (different $\textsf{pk}$, or non-empty $\mathrm{ctx}$) it
        deliberately rejects. Skipping any of the algorithm's
        validation steps (norm bound on $\mathbf{z}$, well-formedness
        of $\mathbf{h}$, monotone hint encoding) is \emph{not}
        permitted within $\mathcal{I}_{\mathrm{sb}}$. Internal
        representations and arithmetic detours that are unobservable
        at the accept/reject boundary (e.g.\ working entirely in
        NTT-Montgomery domain, or precomputing $\NTT(\mathbf{A})$) are
        permitted. The one case in this series where we have not
        closed the equivalence proof even on $\mathcal{I}_{\mathrm{sb}}$
        is $\textsf{v9}$'s skipped pre-invNTT reduction; we flag it
        prominently rather than asserting full conformance there.
  \item \textbf{Empty context.} The bootloader signs without a context
        string ($|\mathrm{ctx}|=0$). This is the only message-shape
        assumption the optimized verifier actually \emph{bakes in}: the
        2-byte $\mathtt{0x00\,0x00}$ prefix becomes part of the
        pre-warmed Keccak template at $\textsf{v5}$, and the verifier
        explicitly rejects inputs with non-zero $|\mathrm{ctx}|$. The
        message itself flows through the verifier with an arbitrary
        length \texttt{mlen} (in our benchmark, $\mathtt{mlen}=32$
        because the bootloader signs over $\mathrm{SHA}\text{-}256$ of
        the image, but no version specializes around the value 32).
\end{enumerate}

We adopt the convention that ``do not cheat'' means: every value the
verifier outputs on inputs in $\mathcal{I}_{\mathrm{sb}}$ is the value
FIPS~204 would compute on the same inputs. Pre-storing
$\NTT(\mathbf{A})$ in flash is legitimate; hardcoding the verify result
is not.

\paragraph{Where precomputes live.} In the bench harness all
precomputed tables live in \texttt{.bss} (RAM-backed
\texttt{static}~arrays) so the QEMU run is self-contained. A real
secure-boot deployment would place them in flash; on a device with
flash wait-states, that changes the read-side cost of every verify by
the wait-state-times-cache-miss-rate factor. Our instruction-retired
metric does not model this, so the absolute cycle counts on real
silicon will be slightly higher than a naive \texttt{insns} $\to$
\texttt{cycles} translation suggests when precomputes are in flash;
the relative ranking of variants is unaffected.

\section{Optimizations}
\label{sec:opts}

We now walk through the ten progressive variants
$\textsf{v1}\ldots\textsf{v10}$. Each subsection states the high-level
intuition (the ``why''), the precise transformation applied (the ``what''),
and the measured impact (the ``so what''). Throughout, the baseline is
$\textsf{v1}$: the unmodified pqm4 \texttt{m4f} verify with secure-boot--shaped
inputs. The cumulative speedup over $\textsf{v1}$ is tracked in
Table~\ref{tab:results} (Section~\ref{sec:results}).

\subsection{v1 --- Baseline}

A direct call to \texttt{pqm4}'s \texttt{crypto\_sign\_verify\_ctx} with
$M = h$ (32 bytes) and $\mathrm{ctx} = \varepsilon$ measures
$\mathbf{1{,}261{,}555}$ instructions retired. This is $\sim$0.7\% below
the same code running on a 62-byte test message because $\mu$'s SHAKE-256
absorbs fewer bytes.

\subsection{v2 --- Precompute \texorpdfstring{$\mathbf{\hat A} \in \Rq^{k\times\ell}$}{A in NTT}}

\paragraph{Why.} ExpandA generates the $k\ell = 16$ polynomials of
$\mathbf{A}$ via SHAKE-128 rejection sampling streamed from
$\rho \,\|\, j \,\|\, i$. Because $\rho$ is the first 32 bytes of
$\textsf{pk}$, and $\textsf{pk}$ is fixed at provisioning, the entire
matrix $\mathbf{A}$ is a constant. Storing it in flash --- as $k \cdot \ell
\cdot n \cdot 4 = 16$\,KB of 32-bit Montgomery-domain NTT coefficients ---
eliminates the dominant cost of verify.

\paragraph{What.} We add a one-shot \texttt{sb\_v2\_precompute(pk)} that
calls \texttt{poly\_uniform} for every $(i,j)$ pair and stores the
result in a static \texttt{poly sb\_A\_NTT[K][L]}. The verify hot loop
reads \texttt{\&sb\_A\_NTT[i][j]} where the baseline called
\texttt{expand\_mat\_elem}.

\paragraph{So what.} $\textsf{v2}$ costs $\mathbf{410{,}292}$ instructions
per verify --- a $\mathbf{3.07\times}$ speedup and a $\mathbf{-67.5\%}$
reduction. The precompute itself costs $859{,}570$ instructions, paid
once.

\subsection{v3 --- Precompute \texorpdfstring{$\NTT(2^d \cdot t_1)$}{NTT(2^d * t1)}}

\paragraph{Why.} The verifier needs $\hat c \cdot \NTT(2^d t_1)$ on every
verify, but only $\hat c$ depends on the signature. The factor
$\NTT(2^d t_1)$ depends only on $\textsf{pk}$ (since $t_1$ is the
$\Power$-truncated component of $\textsf{pk}$) and can be moved to
precompute.

\paragraph{What.} For each row index $k$, $\textsf{v3}$'s precompute does
\texttt{unpack\_pk\_t1} (10-bit unpack of 256 coefficients from the packed
section of $\textsf{pk}$), \texttt{poly\_shiftl} (multiply by $2^{13}$), and
\texttt{poly\_ntt}, storing the result in
\texttt{poly sb\_t1\_ntt[K]}. The verify hot loop replaces the three
operations above with a single pointwise multiplication.

\paragraph{So what.} $\textsf{v3}$ costs $\mathbf{368{,}019}$ instructions
($\mathbf{3.43\times}$ over $\textsf{v1}$; $-10.3\%$ over $\textsf{v2}$).
The 4\,KB precompute is small.

\subsection{v4 --- Precompute \texorpdfstring{$tr = \SHAKE_{256}(\textsf{pk}, 64)$}{tr = SHAKE-256(pk, 64)}}

\paragraph{Why.} The message representative
$\mu = \SHAKE_{256}(tr \,\|\, \mathtt{0x00} \,\|\, \mathrm{IntegerToBytes}(|\mathrm{ctx}|,1) \,\|\, \mathrm{ctx} \,\|\, M, 64)$
is computed in pqm4 by first hashing the whole 1312-byte $\textsf{pk}$
into $tr$, then absorbing $tr$ and the rest into a second SHAKE-256.
Hashing $\textsf{pk}$ with the rate-136 SHAKE-256 absorbs nine full
blocks ($9 \cdot 136 = 1224$ bytes) followed by an 88-byte partial
block; the spec-mandated finalize then runs one more permutation. So
\textbf{nine absorb permutations plus one finalize permutation} are
spent per verify on a quantity that depends only on $\textsf{pk}$.

\paragraph{What.} The precompute stores
\texttt{static uint8\_t sb\_tr\_pk[TRBYTES = 64]}. The verify path absorbs
\texttt{sb\_tr\_pk} directly instead of calling \texttt{shake256(tr, 64, pk, 1312)}.

\paragraph{So what.} $\mathbf{268{,}420}$ instructions ($\mathbf{4.70\times}$;
$-27.1\%$ over $\textsf{v3}$). The 64 bytes of flash buys the 10
Keccak permutations and the surrounding SHAKE-256 plumbing.

\subsection{v5 --- Pre-Warmed Keccak State for \texorpdfstring{$\mu$}{mu}}

\paragraph{Why.} After $\textsf{v4}$, the $\mu$ computation initializes a
SHAKE-256 state, XORs in $tr$ (64 bytes), then XORs in the 2-byte domain
separator $\texttt{(0x00, |}\mathrm{ctx}\texttt{|)}$. Because
$|\mathrm{ctx}| = 0$ is fixed for secure boot, the post-prefix state is
also constant: a 208-byte struct (the 200-byte Keccak state plus an
8-byte position counter holding the value 66).

\paragraph{What.} The precompute walks a SHAKE-256 incremental state up to
having $tr \,\|\, \mathtt{0x00} \,\|\, \mathtt{0x00}$ absorbed, and stores
the resulting struct. Verify \texttt{memcpy}s the template, absorbs the
32-byte message, finalizes, squeezes.

\paragraph{So what.} A near-wash: $\mathbf{267{,}944}$ instructions
($\mathbf{4.71\times}$; $-476$ instructions over $\textsf{v4}$). On
Cortex-M4 the cost of \texttt{memcpy}ing 208~bytes happens to almost
equal the cost of zero-initializing the state and XORing the 66 prefix
bytes. We commit the change for the sake of the abstract precompute
boundary but acknowledge the negligible gain.

\subsection{v6 --- Decode the Hint Vector Once}

\paragraph{Why.} pqm4's \texttt{unpack\_sig\_h(\&h, idx, sig)} extracts
the hint polynomial for one row, but to validate the strong-unforgeability
monotone-index requirement cumulatively it walks all $k$ rows of the
encoded hint block on every call. Calling it $k$ times therefore visits
the encoding $k^2$ times.

\paragraph{What.} A single function
\texttt{sb\_v6\_unpack\_sig\_h\_all(hs, sig)} fills every
\texttt{hs[i]} in one sweep, zero-fills $kn$ coefficients, validates the
encoding once, and stores hint bits at the indicated coefficient
positions.

\paragraph{So what.} $\mathbf{265{,}460}$ instructions
($\mathbf{4.75\times}$; $-2{,}484$). Smaller than expected because
$-$O3 hoists much of the per-call zero-fill check; the wins concentrate
on the cumulative-validation walk.

\subsection{v7 --- Fuse \texttt{caddq + UseHint + polyw1\_pack}}

\paragraph{Why.} The per-row verify tail traverses the 256-coefficient
$\mathbf{w}'$ three times in succession: $\texttt{poly\_caddq}$ to
re-center coefficients into $[0,q)$, $\texttt{poly\_use\_hint}$ to
reconstruct the high bits with $\mathbf{h}$, and
$\texttt{polyw1\_pack}$ to compress to 6-bit-per-coefficient packed
output. Each pass loads and stores 1\,KB of memory; total redundant I/O
per row is $\sim$3\,KB.

\paragraph{What.} A single \texttt{sb\_v7\_caddq\_useHint\_pack(packed, w, h)}
processes coefficients in groups of four (the granularity of the
4-coeffs-in-3-bytes packing layout for $\gamma_2 = (q-1)/88$). Per
coefficient it inlines $\texttt{caddq}$, the integer-arithmetic
$\HighBits$ decomposition
\[
  a_1 \;=\; \big( \lfloor (a + 127)/2^7 \rfloor \cdot 11275 + 2^{23} \big) \gg 24,
  \qquad a_1 \gets 0 \text{ if } a_1 \ge 44,
\]
the hint application $a_1 \gets (a_1 \pm 1) \bmod 44$ on
$\mathbf{h}$-marked positions, and the bit-shifting that packs four
six-bit $a_1$ values into three bytes. The magic constant
$11{,}275 = \lfloor 2^{31}/(2\gamma_2) \rfloor$ with
$2\gamma_2 = (q-1)/44 = 190{,}464$ is a strength-reduced reciprocal
of $2\gamma_2$ scaled to $2^{31}$; combined with the $\gg 7$
preconditioning and the $+2^{23}$ rounding offset before the outer
$\gg 24$, the formula computes round-to-nearest-multiple-of-$2\gamma_2$
(i.e.\ the centred-representative $\HighBits$ as Dilithium defines it,
not a floor) on every $a \in [0, q)$. The final clamp $a_1 \gets 0$ if
$a_1 \ge 44$ folds the wrap-around case for $a$ very close to $q-1$.

\paragraph{So what.} $\mathbf{256{,}559}$ instructions
($\mathbf{4.92\times}$; $-8{,}901$). The savings come from eliminating
intermediate stores rather than from algorithmic change.

\subsection{v8 --- Fuse Unpack and Norm-Check on \texorpdfstring{$\mathbf{z}$}{z}}

\paragraph{Why.} The verifier reads $\mathbf{z}$ from the signature
(18-bit-per-coefficient packed), reconstructs the int32 representation
$z_i = \gamma_1 - \mathrm{raw}_i$, and then re-scans the int32 array to
test $|z_i| < \gamma_1 - \beta$. Two passes over $\ell n = 1024$
coefficients.

\paragraph{What.} The norm check translates cleanly to a check on the raw
18-bit value: $|\gamma_1 - \mathrm{raw}| < \gamma_1 - \beta$ iff
$\beta < \mathrm{raw} < 2^{18} - \beta$. For ML-DSA-44 with $\beta = 78$,
the predicate $\mathrm{raw} \le 78 \;\vee\; \mathrm{raw} \ge 262{,}066$
identifies an out-of-range coefficient. The fused
\texttt{sb\_v8\_unpack\_sig\_z\_chknorm} writes the int32 only after the
predicate passes for all four coefficients of a group.

\paragraph{So what.} $\mathbf{247{,}788}$ instructions
($\mathbf{5.09\times}$; $-8{,}771$). One full pass over 1\,KB of int32
data eliminated.

\paragraph{Aside.} The original v8 also attempted to XOR the
6-bit-packed $\mathbf{w}_1$ bytes directly into the
$\tilde c'$ SHAKE-256 state's rate, eliminating the 192-byte
per-row stack buffer. The implementation produced a different
$\tilde c'$ than the inc-API reference; the bug appears to be a
write-barrier issue between byte writes through a
\texttt{(uint8\_t\,*)} alias and the subsequent
\texttt{KeccakF1600\_StatePermute((uint64\_t\,*))} call, but I was unable
to isolate it inside the time budget. The win is partially recovered in
$\textsf{v10}$.

\subsection{v9 --- Skip \texttt{poly\_reduce} Before \texorpdfstring{$\iNTT$}{invNTT}}

\paragraph{Why.} After pqm4's \texttt{poly\_pointwise\_montgomery} /
\texttt{poly\_pointwise\_acc\_montgomery}, each NTT-Montgomery
coefficient sits in roughly $(-q, q]$. After accumulating $\ell = 4$
such products in the matrix--vector inner loop and subtracting one
more product for $\hat c \cdot \NTT(2^d t_1)$, an a-priori bound is
$|\hat{w}'_i| \le 5q$, i.e.\ comfortably inside 32-bit signed
arithmetic. pqm4 calls \texttt{poly\_reduce} next to compress the range
toward $[-(q-1)/2,(q-1)/2]$ before feeding $\iNTT$; empirically the
\texttt{m4f} invNTT-tomont yields the correct value on the un-reduced
input for the vectors we tested. The reduction is therefore
\emph{conjecturally} redundant on this call site.

\paragraph{What.} Delete the \texttt{poly\_reduce(\&w1)} call.

\paragraph{So what.} $\mathbf{242{,}376}$ instructions
($\mathbf{5.20\times}$; $-5{,}412$). The bench verifies the
secure-boot test vector across every $\textsf{ITERS}$ we run with
$g_{\mathrm{acc}} = 0$ (verifier returns success).

\paragraph{Caveat.} This is the only optimization in the series whose
correctness we have not algebraically established. Specifically, we
do not prove that pqm4's \texttt{m4f} invNTT-tomont produces the
correct output for every input in $[-5q, 5q]^{n}$ --- only that for
this one valid signature it does. A production deployment should
either (a) tighten the bound by adding a worst-case analysis of the
matrix--vector accumulator's Montgomery-domain output, (b) keep
\texttt{poly\_reduce}, or (c) replace the M4 invNTT with one that
documents its admissible input range. We mark this as an open
soundness item rather than a confirmed correctness result.

\subsection{v10 --- Raw Keccak Driving for \texorpdfstring{$\tilde c'$}{c-tilde-prime}}

\paragraph{Why.} The inc-API of \texttt{pqm4/mupq/common/fips202.c} wraps
each Keccak operation with parameter marshalling and a position-tracking
struct (\texttt{uint64\_t ctx[26]}, where \texttt{ctx[25]} holds the
position counter). Each \texttt{shake256\_inc\_absorb} call performs the
position-counter bookkeeping; calling it $1 + k = 5$ times per verify
adds plumbing.

\paragraph{What.} We bypass the inc-API entirely for the
$\tilde c'$ stream and call the underlying asm primitives ---
\texttt{KeccakF1600\_StateXORBytes}, \texttt{KeccakF1600\_StatePermute},
\texttt{KeccakF1600\_StateExtractBytes} --- directly. The position
counter lives in a local \texttt{size\_t pos}, kept in a register across
the loop body.

\paragraph{So what.} $\mathbf{241{,}756}$ instructions
($\mathbf{5.22\times}$; $-621$). The win is small because the inc-API
plumbing wasn't large to begin with --- but it retires the loose end
from $\textsf{v8}$'s failed direct-XOR experiment.

\section{Results}
\label{sec:results}

\subsection{Measurement Methodology}

All numbers are \emph{instructions retired} on the Cortex-M4 ISA as
observed by a QEMU TCG plugin~\cite{cyclebench-m4} running
\texttt{qemu-system-arm 11.0.0 -M mps2-an386}. The benches are bare-metal
ELFs compiled with \texttt{arm-none-eabi-gcc 13.2.1}
(\texttt{-mcpu=cortex-m4 -mthumb -mfpu=fpv4-sp-d16 -mfloat-abi=hard
-O3}). Instructions retired are a \emph{lower bound} on real-silicon
cycles --- they count each retired instruction as one and ignore taken-
branch refill, single-cycle vs.\ multi-cycle \texttt{LDR}/\texttt{STR}
pairing, the 3--7-cycle range of \texttt{SMULL}/\texttt{UMULL}/
\texttt{SMLAL}, the 2--12-cycle \texttt{SDIV}/\texttt{UDIV}, and any
flash wait states the device imposes. The metric is therefore best used
for \emph{comparing} implementations of the same algorithm rather than
for absolute wall-clock prediction: a real-silicon-cycles to
instructions-retired ratio of roughly $1.3\times$--$2.5\times$ is
typical for cryptographic workloads on Cortex-M4, with the higher end
applying to memory-bound code on slow flash and the lower end applying
to register-rich loops in tightly-cached code. We cross-checked one
point against published silicon numbers: pqm4 lists $\sim$1.65M cycles
for ML-DSA-44 verify on STM32F407, vs.\ our $1{,}270{,}759$
instructions retired for the same code, giving a ratio of $\sim$$1.3$.

Per-call cost is extracted as
$(\mathrm{insns}(\textsf{ITERS}=11) - \mathrm{insns}(\textsf{ITERS}=1)) / 10$,
which cancels program startup, the one-shot precompute, and the trailing
semihosting writes. The methodology is bit-exactly reproducible because
QEMU's TCG executes deterministically and our test inputs are
hardcoded; we verified that the formula yields the same answer at
$\textsf{ITERS}=101$.

\subsection{The Table}

\begin{table}[h]
\centering
\caption{Per-verify cost on Cortex-M4, instructions retired. Every
\textsf{sb-mldsa44-v\textit{N}} ($N\ge 2$) precomputes from $\textsf{pk}$
once before the timed loop; the precompute cost is amortized away by the
measurement formula.}
\label{tab:results}
\begin{tabular}{l r r r}
\toprule
Variant                                      & Insns retired & vs $\textsf{v1}$ & $\Delta$ vs prev. \\
\midrule
$\textsf{v1}$ pqm4 m4f baseline               & $1{,}261{,}555$ & $1.00\times$ & --- \\
$\textsf{v2}$ + precompute $\mathbf{\hat A}$  & $410{,}292$     & $3.07\times$ & $-67.5\%$ \\
$\textsf{v3}$ + precompute $\NTT(2^d t_1)$    & $368{,}019$     & $3.43\times$ & $-10.3\%$ \\
$\textsf{v4}$ + precompute $tr_{\textsf{pk}}$ & $268{,}420$     & $4.70\times$ & $-27.1\%$ \\
$\textsf{v5}$ + pre-warmed $\mu$ Keccak       & $267{,}944$     & $4.71\times$ & $-0.2\%$ \\
$\textsf{v6}$ + one-shot hint decode          & $265{,}460$     & $4.75\times$ & $-0.9\%$ \\
$\textsf{v7}$ + caddq/UseHint/pack fusion     & $256{,}559$     & $4.92\times$ & $-3.4\%$ \\
$\textsf{v8}$ + unpack/chknorm fusion         & $247{,}788$     & $5.09\times$ & $-3.4\%$ \\
$\textsf{v9}$ -- redundant \texttt{poly\_reduce} & $242{,}376$  & $5.20\times$ & $-2.2\%$ \\
$\textsf{v10}$ + raw Keccak driver            & $\mathbf{241{,}756}$ & $\mathbf{5.22\times}$ & $-0.3\%$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Library Context}

Placing $\textsf{v10}$ alongside the off-the-shelf verifiers we measured
in earlier work (Table~\ref{tab:cross}):

\begin{table}[h]
\centering
\caption{Cross-library verify cost on Cortex-M4 (instructions retired
under $-$O3 with the harness described in
Section~\ref{sec:results}). The two Dilithium2 entries below the line
follow the round-3 Dilithium specification rather than FIPS 204
ML-DSA-44, so they do not share byte-level test vectors with the
upper entries; the algebra is identical and the instruction counts
are directly comparable.}
\label{tab:cross}
\begin{tabular}{l r r}
\toprule
Implementation                              & Insns retired   & vs \textsf{v10} \\
\midrule
$\textsf{sb-mldsa44-v10}$ (this work)        & $\mathbf{241{,}756}$ & $1.00\times$ \\
pqm4 m4f ML-DSA-44 (generic, M4 asm)         & $1{,}270{,}759$       & $5.26\times$ \\
mldsa-native ML-DSA-44 (portable C90)        & $2{,}120{,}554$       & $8.77\times$ \\
\midrule
FasterKyberDilithiumM4 Dilithium2 ``new''~\cite{fkd} & $1{,}335{,}713$ & $5.52\times$ \\
\midrule
oreparaz/p256 ECDSA (pure C, incl.\ SHA-256) & $8{,}122{,}459$       & $33.6\times$ \\
p256-m ECDSA (M4 UMAAL asm)                  & $12{,}320{,}435$      & $51.0\times$ \\
\bottomrule
\end{tabular}
\end{table}

The FasterKyberDilithiumM4 measurement is worth a remark. The repository's
published wins are concentrated on the signing side (where the rejection
loop dominates and the small-modulus NTTs are most useful); for verify
the structure is essentially pqm4's loop with the matrix materialized
on the stack rather than streamed. The 5\% gap to \texttt{pqm4 m4f} we
measure reflects that materialization plus minor compiler scheduling
differences --- not the underlying primitives, which are largely the
same hand-tuned assembly.

\section{Discussion}
\label{sec:discussion}

\subsection{What This Is and Is Not}

$\textsf{sb-mldsa44-v10}$ is a \emph{deployment-specialized} verifier, not
an algorithmic improvement to ML-DSA. The savings come from amortizing
$\textsf{pk}$-derived computation across boots and from removing
abstraction layers around the underlying primitives. It does \emph{not}
replace or weaken any FIPS-204 step: the spec-mandated norm bound on
$\mathbf{z}$, the strong-unforgeability monotone-encoding check on
$\mathbf{h}$, and the recompute-and-compare of $\tilde c'$ all remain in
place.

It is also not a \emph{constant-time} verifier: the timing of
\texttt{sb\_v8\_unpack\_sig\_z\_chknorm}'s early return, of
\texttt{sb\_v6\_unpack\_sig\_h\_all}'s monotone check, and of the
hint-conditional branch inside \texttt{sb\_v7\_caddq\_useHint\_pack} all
depend on the signature contents. For secure-boot --- where the
signature is public --- this is exactly the right trade.

\subsection{The Long Tail vs.\ the Two Big Wins}

The cumulative speedup ladder in Table~\ref{tab:results} has a
characteristic ``two stairs and a slope'' shape. The two
$\textsf{pk}$-precompute transformations $\textsf{v2}$ (matrix
$\mathbf{A}$, $851{,}263$ insns) and $\textsf{v4}$ ($tr$ hash,
$99{,}599$ insns) together account for $950{,}862$ of the total
$1{,}019{,}799$ instructions removed across all ten variants ---
$93.2\%$ of the cumulative savings, or equivalently $75.4\%$ of
$\textsf{v1}$'s original cost. The remaining seven transformations
($\textsf{v3}$ and $\textsf{v5}$--$\textsf{v10}$) contribute
$68{,}937$ instructions total, or $6.8\%$ of cumulative savings
($5.5\%$ of $\textsf{v1}$). The lesson is that precomputing
$\textsf{pk}$-derived data is the strategically important move; the
rest are tactical fusions and de-abstractions.

\subsection{What's Not Optimized (Limits of This Work)}

We did not write any new Cortex-M4 assembly. The hot polynomial
primitives --- NTT, inverse NTT, pointwise-Mont, Keccak-$f[1600]$ ---
remain pqm4's \texttt{m4f} asm. A plausible next step would be a
custom \texttt{poly\_pointwise\_sub\_montgomery} that folds the
$\hat c \cdot \NTT(2^d t_1)$ pointwise multiplication with the
subsequent \texttt{poly\_sub} from $\hat{\mathbf{w}}$ into a single
fused pass, saving the round-trip through the intermediate
\texttt{tmp\_elem} polynomial. We do not estimate the achievable
saving here.

\subsection{What Remains on the Critical Path}

The $241{,}756$ instructions of $\textsf{v10}$ are dominated by
quantities that genuinely depend on per-verify inputs and that
Table~\ref{tab:cost-breakdown} therefore does not account for: the
four forward NTTs on $\mathbf{z}$, one forward NTT on $c$, the
$k\ell = 16$ pointwise-Montgomery accumulations of the matrix--vector
product, the $k = 4$ inverse NTTs on $\hat{\mathbf{w}}'$, the
$\SHAKE$-256 absorbs that compute $\mu$ from the message, the
$\SHAKE$-256 stream that computes $\tilde c'$ from $\mu$ and
$\mathbf{w}_1$, the \texttt{SampleInBall} of $c$, and the spec-mandated
unpacking and norm/encoding checks. Estimating individual costs further
would require selectively instrumenting each phase, which we have not
done; we therefore present only what we can measure directly.

At a 100\,MHz Cortex-M4 with typical wait-state penalties, $241{,}756$
instructions retired corresponds to roughly $3$--$6$\,ms of wall-clock
time. Comfortably inside the latency budget of any secure-boot
deployment we are aware of.

\section{Reproducibility}

The source tree is at \url{https://github.com/oreparaz/mldsa-bench}. The
QEMU plugin is from \texttt{cyclebench-m4}~\cite{cyclebench-m4}; the
underlying ML-DSA implementations are
\texttt{mldsa-native}~\cite{mldsanative} (the C90 baseline) and
\texttt{pqm4}~\cite{pqm4} (the M4 asm baseline and the substrate for
the secure-boot variants). After \texttt{./setup.sh},
\texttt{(cd cyclebench-m4 \&\& make qemu)}, and
\texttt{make -C cyclebench-m4/plugin}, the sweep is
\texttt{(cd bench \&\& ./run\_sweep.sh)}.

\section{Conclusion}

ML-DSA-44 signature verification on a Cortex-M4 was already faster than
P-256 ECDSA verification in every off-the-shelf configuration we
benchmarked. By exploiting two structural properties of secure-boot
deployments --- a fixed public key and all-public inputs --- we reduced
the cost by an additional $5.22\times$, to $\sim$242\,k instructions
retired per verify. The two $\textsf{pk}$-precomputation steps
($\textsf{v2}$ and $\textsf{v4}$) account for $75.4\%$ of $\textsf{v1}$'s
original cost (or, equivalently, $93.2\%$ of all instructions we removed
across the ten variants); the remaining seven transformations contribute
another $5.5\%$ of original cost. The verifier reproduces the FIPS-204
accept/reject decision on the secure-boot input subset
$\mathcal{I}_{\mathrm{sb}}$ defined in Section~\ref{sec:secure-boot} ---
modulo the open soundness item flagged at $\textsf{v9}$ --- and
deliberately rejects every input outside that subset.

\begin{thebibliography}{9}
\bibitem{dilithium}
L.~Ducas, E.~Kiltz, T.~Lepoint, V.~Lyubashevsky, P.~Schwabe, G.~Seiler,
D.~Stehl\'e.
\newblock CRYSTALS-Dilithium: A Lattice-Based Digital Signature Scheme.
\newblock \emph{IACR Transactions on Cryptographic Hardware and Embedded
Systems}, 2018(1), pp.~238--268.

\bibitem{fips204}
National Institute of Standards and Technology.
\newblock FIPS 204: Module-Lattice-Based Digital Signature Standard.
\newblock 2024. \url{https://csrc.nist.gov/pubs/fips/204/final}.

\bibitem{pqm4}
M.~J.~Kannwischer, P.~Schwabe, D.~Stebila, T.~Wiggers.
\newblock Improving Software Quality in Cryptography Standardization
Projects.
\newblock \emph{IEEE European Symposium on Security and Privacy
Workshops}, 2022, pp.~19--30.
\newblock \url{https://github.com/mupq/pqm4}.

\bibitem{mldsanative}
The Post-Quantum Cryptography Alliance.
\newblock \texttt{mldsa-native}: a portable C90 implementation of
ML-DSA.
\newblock \url{https://github.com/pq-code-package/mldsa-native}.

\bibitem{cyclebench-m4}
O.~Reparaz.
\newblock \texttt{cyclebench-m4}: A small harness for estimating
Cortex-M4 cycle counts via QEMU's TCG plugin API.
\newblock \url{https://github.com/oreparaz/cyclebench-m4}.

\bibitem{p256-m}
M.~P\'egouri\'e-Gonnard.
\newblock \texttt{p256-m}: a minimalistic implementation of ECDH and
ECDSA on NIST P-256.
\newblock \url{https://github.com/mpg/p256-m}.

\bibitem{fkd}
\texttt{FasterKyberDilithiumM4}: a Cortex-M4 implementation of Kyber
and Dilithium with further NTT and pointwise improvements over
\texttt{pqm4}.
\newblock \url{https://github.com/FasterKyberDilithiumM4/FasterKyberDilithiumM4}.

\end{thebibliography}

\end{document}
